---
title: "Investigate pink/white polymorphism in Leptosiphon parviflorus"
output: html_notebook
---

# clean slate

```{r}
rm(list = ls())
gc()
.rs.restartR(clean = TRUE)
```

# set-up

```{r}
# packages
library(ggplot2)
library(tidyverse)
library(data.table)
library(ggforce)
library(ape)
library(scico)
today <- Sys.Date()

# global ggplot2 theme
theme_set(
  theme_minimal() +
    theme(
      text = element_text(
        size = 16,
        family = "Helvetica",
        colour = "black"
      ),
      axis.text = element_text(colour = "black"),
      plot.title = element_text(hjust = 0.5),
      strip.text = element_text(colour = "black"),
      panel.grid.minor = element_blank()
    )
)
```

# sift interproscan output

```{r}

part1 <- fread("../../resources/interproscan/iprscan5-R20260721-003940-0554-50156080-p1m.tsv")

part2 <- fread("../../resources/interproscan/iprscan5-R20260721-003958-0335-42354918-p2m.tsv")

interproscan_results <- rbind(part1, part2)

panther_results <- interproscan_results %>% 
  filter(V4 == "PANTHER") %>%
  filter(!str_detect(V6, "POLYMERASE")) %>%
  filter(!str_detect(V6, "TRANSPOSASE")) %>%
  filter(!str_detect(V6, "DUF"))

```

# dotplots

## haplotype 1 vs primary

```{r}
# ---- 1. Read PAF ----
paf_cols <- c("qname", "qlen", "qstart", "qend", "strand",
              "tname", "tlen", "tstart", "tend",
              "nmatch", "alen", "mapq")

paf <- read_tsv("../../resources/dotplots/hap1_vs_primary.paf", col_names = FALSE, col_types = cols(.default = "c"))
paf <- paf[, 1:12]                      # keep only the 12 mandatory columns
colnames(paf) <- paf_cols

paf <- paf %>%
  mutate(across(c(qlen, qstart, qend, tlen, tstart, tend, nmatch, alen, mapq), as.numeric),
         pident = nmatch / alen)        # approximate % identity per alignment block

# ---- 2. Filter low alignment blocks ----
paf_f <- paf %>%
  filter(alen >= 1000, mapq >= 1)

# ---- 3. Orient reverse-strand alignments for correct diagonal direction ----
paf_plot <- paf_f %>%
  mutate(
    tstart_plot = if_else(strand == "-", tend, tstart),
    tend_plot   = if_else(strand == "-", tstart, tend)
  )

# ---- 4. Simple whole-genome dotplot ----
ggplot(paf_plot) +
  geom_segment(aes(x = qstart, xend = qend,
                    y = tstart_plot, yend = tend_plot,
                    color = strand),
               linewidth = 0.6) +
  scale_color_manual(values = c("+" = "steelblue", "-" = "firebrick")) +
  labs(x = "Query (bp)", y = "Target (bp)",
       title = "minimap2 whole-genome alignment",
       color = "Strand") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())

## 5. specific scaffold
for(tscaf in paste0("scaffold_", 1:51)){
    
  ggplot(subset(paf_plot, qname == "scaffold_25" & tname == tscaf)) +
  geom_segment(aes(x = qstart, xend = qend,
                    y = tstart_plot, yend = tend_plot,
                    color = strand),
               linewidth = 0.6) +
  scale_color_manual(values = c("+" = "steelblue", "-" = "firebrick")) +
  labs(x ="Haplotype 1, scaffold 25 (bp)", y = paste0("Primary ", tscaf, " (bp)"),
       title = "minimap2 whole-genome alignment",
       color = "Strand") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
  
  ggsave(paste0("../../results/by-date/", today, "/dotplot_hap1_scaffold_25_vs_primary_scaffold_", tscaf, ".pdf"), width = 6, height = 6)
}

```

## haplotype 1 and haplotype 2

check if myb cluster also present on haplotype 2

```{r}
paf_cols <- c("qname", "qlen", "qstart", "qend", "strand",
              "tname", "tlen", "tstart", "tend",
              "nmatch", "alen", "mapq")

paf <- read_tsv("../../resources/dotplots/hap2_vs_hap1.paf", col_names = FALSE, col_types = cols(.default = "c"))
paf <- paf[, 1:12]                      # keep only the 12 mandatory columns
colnames(paf) <- paf_cols

paf <- paf %>%
  mutate(across(c(qlen, qstart, qend, tlen, tstart, tend, nmatch, alen, mapq), as.numeric),
         pident = nmatch / alen)        # approximate % identity per alignment block

# ---- 2. Filter low alignment blocks ----
paf_f <- paf %>%
  filter(alen >= 1000, mapq >= 1)

# ---- 3. Orient reverse-strand alignments for correct diagonal direction ----
paf_plot <- paf_f %>%
  mutate(
    tstart_plot = if_else(strand == "-", tend, tstart),
    tend_plot   = if_else(strand == "-", tstart, tend)
  )

# ---- 4. Simple whole-genome dotplot ----
ggplot(subset(paf_plot, tname == "scaffold_25" & qname == "scaffold_2")) +
  geom_segment(aes(x = qstart, xend = qend,
                    y = tstart_plot, yend = tend_plot,
                    color = strand),
               linewidth = 0.6) +
  scale_color_manual(values = c("+" = "steelblue", "-" = "firebrick")) +
  geom_hline(yintercept = 10964369, lty = 1, color = "pink") +
  geom_hline(yintercept = 12046688, lty = 1, color = "pink") +
  annotate("point", x = (258876087+258880517)/2, y = (10964369+12046688)/2, color = "grey", size = 0.5, shape = 17) +
  labs(x = "White (i.e. Haplotype 2), scaffold 2 (bp)", y = "Pink (i.e. Haplotype 1), scaffold 25 (bp)",
       title = "minimap2 white vs pink alignment\nMYB cluster delinated with pink lines",
       color = "Strand") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())

ggsave(paste0("../../results/by-date/", today, "/dotplot_hap2_scaffold2_vs_hap1_scaffold_25.pdf"), width = 6, height = 6)


```
## haplotype 1 vs haplotype 1

Given presence/absence variant of MYB cluster, maybe some scaffolds are not correctly incorporated into the largest 18 scaffolds

```{r}
paf_cols <- c("qname", "qlen", "qstart", "qend", "strand",
              "tname", "tlen", "tstart", "tend",
              "nmatch", "alen", "mapq")

paf <- read_tsv("../../resources/dotplots/hap1_vs_hap1.paf", col_names = FALSE, col_types = cols(.default = "c"))
paf <- paf[, 1:12]                      # keep only the 12 mandatory columns
colnames(paf) <- paf_cols

paf <- paf %>%
  mutate(across(c(qlen, qstart, qend, tlen, tstart, tend, nmatch, alen, mapq), as.numeric),
         pident = nmatch / alen)        # approximate % identity per alignment block

# ---- 2. Filter low alignment blocks ----
paf_f <- paf %>%
  filter(alen >= 1000, mapq >= 1)

# ---- 3. Orient reverse-strand alignments for correct diagonal direction ----
paf_plot <- paf_f %>%
  mutate(
    tstart_plot = if_else(strand == "-", tend, tstart),
    tend_plot   = if_else(strand == "-", tstart, tend)
  )

# ---- 4. Simple whole-genome dotplot ----
ggplot(subset(paf_plot, tname == "scaffold_25" & tname != qname)) +
  geom_segment(aes(x = qstart, xend = qend,
                    y = tstart_plot, yend = tend_plot,
                    color = strand),
               linewidth = 0.6) +
  scale_color_manual(values = c("+" = "steelblue", "-" = "firebrick")) +
  geom_hline(yintercept = 10964369) +
  geom_hline(yintercept = 12046688) +
  labs(x = "Query sequence (bp)", y = "Pink (i.e. Haplotype 1), scaffold 25 (bp)",
       title = "minimap2 whole-genome alignment of Pink vs Pink\nMyb cluster delinated with lines",
       color = "Strand") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())

ggsave(paste0("../../results/by-date/", today, "/dotplot_hap2_scaffold2_vs_hap1_scaffold_25.pdf"), width = 6, height = 6)
```

## haplotype 1 and 2

Code and alignment from yaniv

```{r}
x <- read_tsv("../../resources/dotplots/hap1_vs_hap2.1to1.coords", col_names = c("S1","E1" ,"S2","E2","LEN_1","LEN_2","PCT_IDY","LEN_R","LEN_Q","COV_R","COV_Q","REF_TAG","QUERY_TAG"))

make_plot <- function(scaffold){
  queery <- x|>
    filter(REF_TAG == scaffold) |>
    group_by(QUERY_TAG) |>
    summarise(tot = sum(LEN_2))|>
    mutate(prop = tot / sum(tot))|>
    filter(prop > 0.02)|>
    pull(QUERY_TAG)
  ggplot(x|> filter(REF_TAG==scaffold,QUERY_TAG %in% queery),
         aes(x = S1, xend = E1, y = S2, yend = E2 , color = QUERY_TAG))+
    geom_segment(linewidth = 2)+
    geom_vline(xintercept = 57821464) + 
    geom_vline(xintercept = 57828110) + 
    theme_minimal()+
    theme(axis.line = element_line(color = "black"))+
    ggtitle(scaffold)
}

make_plot("scaffold_1")
ggsave(paste0("../../results/by-date/", today, "/dotplot_scaffold1_g8317.pdf"))

scaffolds <- paste0("scaffold_", 1:36)

pdf(
  paste0("../../results/by-date/", today, "/hap1_vs_hap2_scaffolds_1_to_18.pdf"),
  width = 10,
  height = 7
)

purrr::walk(scaffolds, \(scaffold) {
  print(make_plot(scaffold))
})

dev.off()
```

# myb orthogroup gene tree

```{r}
# Read the tree
my_tree <- read.tree("../OG0001281_with_g166640.tree")

# Basic plot
plot(my_tree, 
     type = "phylogram")
```

# mapping bias

```{r}
idxstats <- list.files(path = "../results/vg/hap1_vs_hap2", pattern = "idxstats.txt", recursive = T, full.names = T)

# Read and bind them while keeping the file names
idxstats <- idxstats %>% 
  set_names() %>% 
    map_dfr(fread, .id = "source_file")

# calculate 
idxstats <- idxstats %>%
  group_by(source_file) %>%
  reframe(mapped = sum(V3),
            unmapped = sum(V4)) %>%
  summarize(sample = 1:8,
            group = sample %% 2,
            pct_mapped = mapped/(mapped+unmapped))
```

# genome scans: pi, fst, tajima's D

```{r}
div_scans <- list.files(path = "../results/vg/hap2_vs_hap1", pattern = "interval_diversity", recursive = T, full.names = T)

# Read and bind them while keeping the file names
div_scans <- div_scans %>% 
  set_names() %>% 
  map_dfr(~fread(.x, col.names = c("chrom","start","end","total.masked","total.missing","total.empty","total.numeric","total.invariant","total.passed","missing","numeric","passed","theta_pi","theta_watterson","tajimas_d")), .id = "source_file")

# separate out sample and site class
div_scans$sample <- gsub("../results/vg/hap2_vs_hap1/|/diversity.*", "", div_scans$source_file)

div_scans$site_class <- gsub("../results/vg/hap2_vs_hap1/.*/diversity/|/grenedalf_results_interval.*", "", div_scans$source_file)

head(div_scans)

table(div_scans$site_class, div_scans$sample)

### rename samples
sample_lookup <- c(
  "1" = "1P", "2" = "1W",
  "3" = "2P", "4" = "2W",
  "5" = "3P", "6" = "3W",
  "7" = "4P", "8" = "4W"
)

div_scans <- div_scans %>%
  mutate(sample = recode(sample, !!!sample_lookup))

### include fst
fst_scans <- list.files(path = "../results/vg/hap2_vs_hap1", pattern = "interval_fst", recursive = T, full.names = T)

# Read and bind them while keeping the file names
fst_scans <- fst_scans %>% 
  set_names() %>% 
  map_dfr(~fread(.x, col.names = c("chrom","start","end","total.masked","total.missing","total.empty","total.numeric","total.invariant","total.passed","1:2.missing","1:2.numeric","1:2.passed","1:2.fst","1:3.missing","1:3.numeric","1:3.passed","1:3.fst","1:4.missing","1:4.numeric","1:4.passed","1:4.fst","1:5.missing","1:5.numeric","1:5.passed","1:5.fst","1:6.missing","1:6.numeric","1:6.passed","1:6.fst","1:7.missing","1:7.numeric","1:7.passed","1:7.fst","1:8.missing","1:8.numeric","1:8.passed","1:8.fst","2:3.missing","2:3.numeric","2:3.passed","2:3.fst","2:4.missing","2:4.numeric","2:4.passed","2:4.fst","2:5.missing","2:5.numeric","2:5.passed","2:5.fst","2:6.missing","2:6.numeric","2:6.passed","2:6.fst","2:7.missing","2:7.numeric","2:7.passed","2:7.fst","2:8.missing","2:8.numeric","2:8.passed","2:8.fst","3:4.missing","3:4.numeric","3:4.passed","3:4.fst","3:5.missing","3:5.numeric","3:5.passed","3:5.fst","3:6.missing","3:6.numeric","3:6.passed","3:6.fst","3:7.missing","3:7.numeric","3:7.passed","3:7.fst","3:8.missing","3:8.numeric","3:8.passed","3:8.fst","4:5.missing","4:5.numeric","4:5.passed","4:5.fst","4:6.missing","4:6.numeric","4:6.passed","4:6.fst","4:7.missing","4:7.numeric","4:7.passed","4:7.fst","4:8.missing","4:8.numeric","4:8.passed","4:8.fst","5:6.missing","5:6.numeric","5:6.passed","5:6.fst","5:7.missing","5:7.numeric","5:7.passed","5:7.fst","5:8.missing","5:8.numeric","5:8.passed","5:8.fst","6:7.missing","6:7.numeric","6:7.passed","6:7.fst","6:8.missing","6:8.numeric","6:8.passed","6:8.fst","7:8.missing","7:8.numeric","7:8.passed","7:8.fst")), .id = "source_file")

fst_scans$site_class <- gsub(".*fst/|/grenedalf.*", "", fst_scans$source_file)

div_scans <- div_scans %>% 
  left_join(fst_scans, by = c("chrom", "start", "end", "site_class"))

#### plot total diversity ####
#cds_scans <- subset(div_scans, site_class == "cds")

# 1. Split the data frame into a list of data frames by your variable
data_list <- split(div_scans, list(div_scans$chrom, div_scans$sample, div_scans$site_class), drop = T)

# 2. Open the PDF device
pdf(paste0("../../results/by-date/", today, "/nucleotide_diversity_scans.pdf"), width = 8, height = 6)

# 3. Map over the list, generate a standard plot for each subset, and print
walk(data_list, function(sub_data) {
p <- ggplot(sub_data,  aes(x = (start + end)/2, y = theta_pi)) +
  geom_point(shape = 16) +
  geom_smooth(method = "loess", span = 0.2, se = FALSE) +
  labs(x = "Window midpoint (bp)", 
       y = "Nucleotide diversity", 
       title = paste0(unique(sub_data$chrom), ", Sample: ", unique(sub_data$sample), ", Sites: ", unique(sub_data$site_class)))
  
  print(p)
})

# 4. Close the PDF device
dev.off()

#### plot tajima's D ####

# 2. Open the PDF device
pdf(paste0("../../results/by-date/", today, "/tajimas_d_scans.pdf"), width = 8, height = 6)

# 3. Map over the list, generate a standard plot for each subset, and print
walk(data_list, function(sub_data) {
p <- ggplot(sub_data,  aes(x = (start + end)/2, y = tajimas_d)) +
  geom_point(shape = 16) +
  geom_smooth(method = "loess", span = 0.2, se = FALSE) +
  labs(x = "Window midpoint (bp)", 
       y = "Tajima's D", 
       title = paste0(unique(sub_data$chrom), 
                      ", Sample: ", 
                      unique(sub_data$sample),
                      ", Sites: ", 
                      unique(sub_data$site_class))
  )

  print(p)
})

# 4. Close the PDF device
dev.off()

### plot piN/piS
div_wide <- dcast(div_scans, chrom + start + end + sample ~ site_class, value.var = "theta_pi")

div_wide$piNpiS <- div_wide$zero/div_wide$four

data_list <- split(div_wide, list(div_wide$chrom, div_wide$sample), drop = T)

# 2. Open the PDF device
pdf(paste0("../../results/by-date/", today, "/piN_piS_scans.pdf"), width = 8, height = 6)

# 3. Map over the list, generate a standard plot for each subset, and print
walk(data_list, function(sub_data) {
p <- ggplot(sub_data,  aes(x = (start + end)/2, y = piNpiS)) +
  geom_point(shape = 16) +
  geom_smooth(method = "loess", span = 0.2, se = FALSE) +
  labs(x = "Window midpoint (bp)", 
       y = "piN/piS", 
       title = paste0(unique(sub_data$chrom), 
                      ", Sample: ", 
                      unique(sub_data$sample)))
  
  print(p)
})

# 4. Close the PDF device
dev.off()

### plot fst

fst_long <- fst_scans %>%
  pivot_longer(
    cols = matches("^\\d+:\\d+\\."),
    names_to = c("pair", ".value"),
    names_pattern = "^(\\d+:\\d+)\\.(.*)$"
  ) %>%
  separate(pair, into = c("idx1", "idx2"), sep = ":", remove = FALSE) %>%
  mutate(
    sample1 = sample_lookup[idx1],
    sample2 = sample_lookup[idx2],
    pair = paste0(sample1, ":", sample2)
  ) %>%
  select(-idx1, -idx2)

fst_long$site_class <- gsub(".*fst/|/grenedalf.*", "", fst_long$source_file)

data_list <- split(fst_long, list(fst_long$chrom, fst_long$pair, fst_long$site_class), drop = T)

# 2. Open the PDF device
pdf(paste0("../../results/by-date/", today, "/fst_scans.pdf"), width = 8, height = 6)

# 3. Map over the list, generate a standard plot for each subset, and print
walk(data_list, function(sub_data) {
  p <- ggplot(sub_data,  aes(x = (start + end)/2, y = fst)) +
    geom_point(shape = 16) +
    geom_smooth(method = "loess", span = 0.2, se = FALSE) +
    labs(x = "Window midpoint (bp)", 
         y = "Fst (unbiased Hudson estimator)", 
         title = paste0(unique(sub_data$chrom), 
                        ", Pair: ", 
                        unique(sub_data$pair),
                        ", Sites:",
                        (sub_data$site_class))
         )
  
  print(p)
})

# 4. Close the PDF device
dev.off()

#### plot only MYB cluster diversity ####

hap1_scaffold_25_div <- div_scans %>%
  filter(chrom == "hap1_scaffold_25",
         site_class == "four")

# 1. Split the data frame into a list of data frames by your variable
data_list <- split(hap1_scaffold_25_div, list(hap1_scaffold_25_div$sample), drop = T)

# 2. Open the PDF device
pdf(paste0("../../results/by-date/", today, "/hap1_scaffold_25_diversity_scans.pdf"), width = 8, height = 6)

# 3. Map over the list, generate a standard plot for each subset, and print
walk(data_list, function(sub_data) {
  
p1 <- ggplot(sub_data,  aes(x = (start + end)/2, y = tajimas_d)) +
  geom_point(shape = 16) +
  geom_smooth(method = "loess", span = 0.2, se = FALSE, color = "grey") +
  geom_vline(xintercept = 10964369, lty = 2, color = "pink", lwd = 1.5) +
  geom_vline(xintercept = 12046688, lty = 2, color = "pink", lwd = 1.5) +
  labs(x = "Window midpoint (bp)", 
       y = "Tajima's D", 
       title = paste0(unique(sub_data$chrom), 
                      ", Sample: ", 
                      unique(sub_data$sample),
                      ", Sites: ", 
                      unique(sub_data$site_class))
  )

  print(p1)
  
  p2 <- ggplot(sub_data,  aes(x = (start + end)/2, y = theta_pi)) +
  geom_point(shape = 16) +
  geom_smooth(method = "loess", span = 0.2, se = FALSE, color = "grey") +
  geom_vline(xintercept = 10964369, lty = 2, color = "pink", lwd = 1.5) +
  geom_vline(xintercept = 12046688, lty = 2, color = "pink", lwd = 1.5) +
  labs(x = "Window midpoint (bp)", 
       y = "Nucleotide diversity", 
       title = paste0(unique(sub_data$chrom), 
                      ", Sample: ", 
                      unique(sub_data$sample),
                      ", Sites: ", 
                      unique(sub_data$site_class))
  )

  print(p2)
})

# 4. Close the PDF device
dev.off()

#### plot only myb cluster fst ####
hap1_scaffold_25_fst <- fst_long %>%
  filter(chrom == "hap1_scaffold_25",
         site_class == "four")

data_list <- split(hap1_scaffold_25_fst, list(hap1_scaffold_25_fst$chrom, hap1_scaffold_25_fst$pair, hap1_scaffold_25_fst$site_class), drop = T)

# 2. Open the PDF device
pdf(paste0("../../results/by-date/", today, "/hap1_scaffold_25_fst_scans.pdf"), width = 8, height = 6)

# 3. Map over the list, generate a standard plot for each subset, and print
walk(data_list, function(sub_data) {
  p <- ggplot(sub_data,  aes(x = (start + end)/2, y = fst)) +
    geom_point(shape = 16) +
    geom_smooth(method = "loess", span = 0.2, se = FALSE, color = "grey") +
  geom_vline(xintercept = 10964369, lty = 2, color = "pink", lwd = 1.5) +
  geom_vline(xintercept = 12046688, lty = 2, color = "pink", lwd = 1.5) +
    labs(x = "Window midpoint (bp)", 
         y = "Fst (unbiased Hudson estimator)", 
         title = paste0(unique(sub_data$chrom), 
                        ", Pair: ", 
                        unique(sub_data$pair),
                        ", Sites:",
                        (sub_data$site_class))
         )
  
  print(p)
})

# 4. Close the PDF device
dev.off()

### plot pi in pink vs pi in white

```

# differentiation matrix

```{r}
dxy4 <- fread("../results/vg/hap2_vs_hap1/fst/four/grenedalf_results_genome_pi-between.csv")

pi4 <- list.files(path = "../results/vg/hap2_vs_hap1", pattern = "grenedalf_results_genome_diversity.csv", recursive = T, full.names = T)

pi4 <- pi4 %>% 
  set_names() %>% 
  map_dfr(~fread(.x, col.names = c("chrom","start","end","total.masked","total.missing","total.empty","total.numeric","total.invariant","total.passed","missing","numeric","passed","theta_pi","theta_watterson","tajimas_d")), .id = "source_file")

pi4$sample_id <- gsub("/", "", str_extract(pi4$source_file, "/[1-8]/"))

pi4$site_class <- gsub("diversity/|/grenedalf", "", str_extract(pi4$source_file, "diversity/.*/grenedalf"))

pi4 <- pi4 %>% filter(site_class == "four")

# reformat grenedalf table into matrix
pi_long <- dxy4 %>%
  select(matches("^\\d+:\\d+\\.pi-between$")) %>%
  pivot_longer(everything(), names_to = "pair", values_to = "pi") %>%
  mutate(pair = str_remove(pair, "\\.pi-between$")) %>%
  separate(pair, into = c("i", "j"), sep = ":", convert = TRUE)

n <- max(pi_long$i, pi_long$j)

pi_mat <- matrix(NA_real_, n, n, dimnames = list(1:n, 1:n))
pi_mat[cbind(pi_long$i, pi_long$j)] <- pi_long$pi
pi_mat[cbind(pi_long$j, pi_long$i)] <- pi_long$pi

diag(pi_mat) <- pi4$theta_pi

rownames(pi_mat) = c("1P", "1W", "2P", "2W", "3P", "3W", "4P", "4W")

colnames(pi_mat) = c("1P", "1W", "2P", "2W", "3P", "3W", "4P", "4W")

# diversity heatmap
# dxy4/pi-between off diagonal
# pi4 on diagonal
pi_df <- pi_mat %>%
  as.data.frame() %>%
  rownames_to_column("row") %>%
  pivot_longer(-row, names_to = "col", values_to = "pi") %>%
  mutate(
    row   = factor(row, levels = rownames(pi_mat)),
    col   = factor(col, levels = colnames(pi_mat)),
    label = round(pi, 4)
  ) %>%
  filter(as.integer(row) >= as.integer(col))

ggplot(pi_df, aes(x = col, y = row, fill = pi)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = label), size = 3.2) +
  scale_fill_scico(palette = "lajolla", name = "\u03c04", direction = -1, begin = 0.25) +
  scale_y_discrete(limits = rev(levels(pi_df$row))) +
  coord_fixed() +
  theme(
    axis.title = element_blank(),
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

ggsave(paste0("../../results/by-date/", today, "/pi4_dxy4_heatmap.pdf"), device = cairo_pdf, width = 6, height = 6)
ggsave(paste0("../../results/by-date/", today, "/pi4_dxy4_heatmap.png"), width = 6, height = 6)

# fst matrix
fst_mat <- fread("../results/vg/hap2_vs_hap1/fst/four/grenedalf_results_genome_fst-matrix.csv")

fst_mat$fst = c("1P", "1W", "2P", "2W", "3P", "3W", "4P", "4W")
colnames(fst_mat) = c("fst", "1P", "1W", "2P", "2W", "3P", "3W", "4P", "4W")

fst_df <- fst_mat %>% 
  pivot_longer(-fst, names_to = "col", values_to = "Fst") %>%
  mutate(
    row   = factor(fst, levels = fst_mat$fst),
    col   = factor(col, levels = colnames(fst_mat)),
    label = round(Fst, 3)
  ) %>%
  filter(as.integer(row) >= as.integer(col))

ggplot(fst_df, aes(x = col, y = row, fill = Fst)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = label), size = 3.2) +
  scale_fill_scico(palette = "buda", name = "Fst", direction = -1, begin = 0.1) +
  scale_y_discrete(limits = rev(levels(pi_df$row))) +
  coord_fixed() +
  theme(
    axis.title = element_blank(),
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

ggsave(paste0("../../results/by-date/", today, "/fst_heatmap.pdf"), device = cairo_pdf, height = 6, width = 6)
ggsave(paste0("../../results/by-date/", today, "/fst_heatmap.png"), height = 6, width = 6)
```

# coverage from mpileup

```{r}

mpileup_files <- list.files(path = "../results/vg/hap2_vs_hap1", pattern = "\\.mpileup$", recursive = T, full.names = T)

mpileup <- mpileup_files %>%
  set_names() %>%
  map_dfr(~fread(
    #cmd = paste0("awk '(($2 < 13129007  && $2 > 9882050))' ", .x, " | cut -f 2,4 --delimiter='\t'"),
    cmd = paste0("awk '(($2 < 12046688  && $2 > 10964369))' ", .x, " | cut -f 2,4 --delimiter='\t'"),
    col.names = c("pos", "depth")), 
    .id = "source_file")

mpileup$sample_id <- gsub("/", "", str_extract(mpileup$source_file, "/[1-8]/"))

total_depths <- mpileup %>%
  group_by(sample_id) %>%
  mutate(total_depth = sum(depth),
            norm_depth = depth/total_depth)


# Recode sample_id
sample_map <- c(
  "1" = "1P", "2" = "1W",
  "3" = "2P", "4" = "2W",
  "5" = "3P", "6" = "3W",
  "7" = "4P", "8" = "4W"
)

total_depths <- total_depths %>%
  mutate(sample_id = recode(sample_id, !!!sample_map))

# Split into pair number + type (P/W), then pivot wider on norm_depth
diff_df <- total_depths %>%
  ungroup() %>%
  mutate(
    pair = substr(sample_id, 1, nchar(sample_id) - 1),  # e.g. "1", "2", "3", "4"
    type = substr(sample_id, nchar(sample_id), nchar(sample_id))  # "P" or "W"
  ) %>%
  select(pos, pair, type, norm_depth) %>%
  pivot_wider(names_from = type, values_from = norm_depth) %>%
  mutate(norm_depth_diff = W - P)

# compare coverage inside vs outside the k-mer mapping region
mpileup %>%
  group_by(sample_id) %>%
  filter(pos > 10964369 & pos < 12046688) %>%
  summarize(P_locus_depth = sum(depth)) %>%
  left_join(total_depths, by = "sample_id") %>%
  mutate(P_locus_lambda = P_locus_depth/total_depth)

mpileup %>%
  group_by(sample_id) %>%
  filter(pos < 13129007 & pos > 12046688) %>%
  summarize(downstream_locus_depth = sum(depth))  %>%
  left_join(total_depths, by = "sample_id") %>%
  mutate(downstream_locus_lambda = downstream_locus_depth/total_depth)

mpileup %>%
  group_by(sample_id) %>%
  filter(pos < 10964369 & pos > 9882050) %>%
  summarize(upstream_locus_depth = sum(depth))  %>%
  left_join(total_depths, by = "sample_id") %>%
  mutate(upstream_locus_lambda = upstream_locus_depth/total_depth)

# coverage scan along the scaffold
ggplot(diff_df %>% slice_sample(prop = 0.01), aes(x = pos, y = norm_depth_diff*1e6)) +
  geom_point(shape = 16) +
  geom_smooth(method = "loess", color = "lightgrey", se = FALSE) +
  geom_vline(xintercept = c(10964369, 12046688), color = "pink") +
  facet_wrap(~pair, nrow = 1, ncol = 4) +
  labs(x = "Position on hap1_scaffold_25 (bp)",
       y = "\u0394 Standardized depth (W-P)",
       title = "Coverage along hap1_scaffold_25") +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

ggsave(paste0("../../results/by-date/", today, "/mpileup_delta_coverage_scan.pdf"), device = cairo_pdf, width = 8, height = 5)
ggsave(paste0("../../results/by-date/", today, "/mpileup_delta_coverage_scan.png"), width = 8, height = 5)

# coverage scan along the scaffold
ggplot(mpileup %>% slice_sample(prop = 0.01), aes(x = pos, y = depth)) +
  geom_point(shape = 16) +
  geom_smooth(method = "loess", se = F, span = 0.2) +
  #geom_vline(xintercept = c(10964369, 12046688)) +
  geom_vline(xintercept = c(11818528, 11822704), lty = 2) +
  facet_wrap(~sample_id, nrow = 2, ncol = 4) +
  labs(x = "Position on hap1_scaffold_25 (bp)",
       y = "Read depth",
       title = "Coverage along hap1_scaffold_25")

ggsave(paste0("../../results/by-date/", today, "/mpileup_coverage_scan.pdf"), device = cairo_pdf, width = 8, height = 6)
ggsave(paste0("../../results/by-date/", today, "/mpileup_coverage_scan.png"), width = 8, height = 6)

# depth distribution per sample
ggplot(mpileup, aes(x = depth, fill = sample_id)) +
  geom_histogram(bins = 50) +
  facet_wrap(~sample_id, scales = "free_y") +
  labs(x = "Read depth",
       y = "Number of positions",
       title = "Read depth distribution on hap1_scaffold_25") +
  theme(legend.position = "none")

ggsave(paste0("../../results/by-date/", today, "/mpileup_depth_histogram.pdf"), device = cairo_pdf, width = 8, height = 6)
ggsave(paste0("../../results/by-date/", today, "/mpileup_depth_histogram.png"), width = 8, height = 6)
```

# k-mer histogram

```{r}

kmc_hist_hap1 <- read.table("/global/scratch/users/milesroberts/brandvain_lab_projects/lepto/workflow/hap1.33.hist", header=F)
kmc_hist_hap2 <- read.table("/global/scratch/users/milesroberts/brandvain_lab_projects/lepto/workflow/hap2.33.hist", header=F)

kmc_hist_hap1$ref <- "hap1"
kmc_hist_hap2$ref <- "hap2"

kmc_hist <- rbind(kmc_hist_hap1, kmc_hist_hap2)

sum(kmc_hist$V2[1])/sum(kmc_hist$V2)

sum(kmc_hist_hap2$V2[1])/sum(kmc_hist_hap2$V2)

ggplot(kmc_hist, aes(x = V1, y = V2, color = ref)) +
  geom_point(alpha = 0.4) +
  geom_line(alpha = 0.4) +
  scale_color_scico_d(palette = "acton", direction = -1, begin = 0.05, end = 0.95) +
  scale_y_continuous(labels = scales::comma) + # Cleans up large scientific notation numbers
  labs(
    title = "Reference k-mer histogram (86% of 33-mers are unique)",
    x = "K-mer count",
    y = "Number of k-mers with count",
    color = "Haplotype"
  ) +
  scale_y_log10() +
  scale_x_log10() +
  theme_minimal() +
  theme(text = element_text(size = 16))

ggsave(paste0("../../results/by-date/", today, "/reference_kmer_histogram.pdf"), device = cairo_pdf, width = 8, height = 6)
ggsave(paste0("../../results/by-date/", today, "/reference_kmer_histogram.png"), width = 8, height = 6)

```

# MSMC2

```{r}

mu <- 7e-9   # substitute species-appropriate value
gen <- 1       # substitute species-appropriate value
dat <- read.table("/global/scratch/users/milesroberts/brandvain_lab_projects/lepto/workflow/out_prefix.final.txt", header=TRUE)


plot(dat$left_time_boundary/mu*gen, (1/dat$lambda)/(2*mu), log="x", type="s",
     xlab="Years ago", ylab="Effective population size", log="y")

```