---
title: "Investigate pink/white polymorphism in Leptosiphon parviflorus"
output: html_notebook
---

# clean slate

```{r}
rm(list = ls())
gc()
.rs.restartR(clean = TRUE)
```

# packages

```{r}
library(ggplot2)
library(tidyverse)
library(data.table)
library(ggforce)
library(ape)
library(scico)
today <- Sys.Date()
```

# global ggplot2 theme

```{r}
theme_set(
  theme_minimal() +
    theme(
      text = element_text(
        size = 16,
        family = "Helvetica",
        colour = "black"
      ),
      axis.text = element_text(colour = "black"),
      plot.title = element_text(hjust = 0.5),
      strip.text = element_text(colour = "black"),
      panel.grid.minor = element_blank()
    )
)
```

# sift interproscan output

```{r}

part1 <- fread("../../resources/interproscan/iprscan5-R20260721-003940-0554-50156080-p1m.tsv")

part2 <- fread("../../resources/interproscan/iprscan5-R20260721-003958-0335-42354918-p2m.tsv")

interproscan_results <- rbind(part1, part2)

panther_results <- interproscan_results %>% 
  filter(V4 == "PANTHER") %>%
  filter(!str_detect(V6, "POLYMERASE")) %>%
  filter(!str_detect(V6, "TRANSPOSASE")) %>%
  filter(!str_detect(V6, "DUF"))

```

# dotplots

## haplotype 1 vs primary

```{r}
# ---- 1. Read PAF ----
paf_cols <- c("qname", "qlen", "qstart", "qend", "strand",
              "tname", "tlen", "tstart", "tend",
              "nmatch", "alen", "mapq")

paf <- read_tsv("../../resources/dotplots/hap1_vs_primary.paf", col_names = FALSE, col_types = cols(.default = "c"))
paf <- paf[, 1:12]                      # keep only the 12 mandatory columns
colnames(paf) <- paf_cols

paf <- paf %>%
  mutate(across(c(qlen, qstart, qend, tlen, tstart, tend, nmatch, alen, mapq), as.numeric),
         pident = nmatch / alen)        # approximate % identity per alignment block

# ---- 2. Filter low alignment blocks ----
paf_f <- paf %>%
  filter(alen >= 1000, mapq >= 1)

# ---- 3. Orient reverse-strand alignments for correct diagonal direction ----
paf_plot <- paf_f %>%
  mutate(
    tstart_plot = if_else(strand == "-", tend, tstart),
    tend_plot   = if_else(strand == "-", tstart, tend)
  )

# ---- 4. Simple whole-genome dotplot ----
ggplot(paf_plot) +
  geom_segment(aes(x = qstart, xend = qend,
                    y = tstart_plot, yend = tend_plot,
                    color = strand),
               linewidth = 0.6) +
  scale_color_manual(values = c("+" = "steelblue", "-" = "firebrick")) +
  labs(x = "Query (bp)", y = "Target (bp)",
       title = "minimap2 whole-genome alignment",
       color = "Strand") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())

## 5. specific scaffold
for(tscaf in paste0("scaffold_", 1:51)){
    
  ggplot(subset(paf_plot, qname == "scaffold_25" & tname == tscaf)) +
  geom_segment(aes(x = qstart, xend = qend,
                    y = tstart_plot, yend = tend_plot,
                    color = strand),
               linewidth = 0.6) +
  scale_color_manual(values = c("+" = "steelblue", "-" = "firebrick")) +
  labs(x ="Haplotype 1, scaffold 25 (bp)", y = paste0("Primary ", tscaf, " (bp)"),
       title = "minimap2 whole-genome alignment",
       color = "Strand") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
  
  ggsave(paste0("../../results/by-date/", today, "/dotplot_hap1_scaffold_25_vs_primary_scaffold_", tscaf, ".pdf"), width = 6, height = 6)
}

```

## haplotype 1 and haplotype 2

check if myb cluster also present on haplotype 2

```{r}
paf_cols <- c("qname", "qlen", "qstart", "qend", "strand",
              "tname", "tlen", "tstart", "tend",
              "nmatch", "alen", "mapq")

paf <- read_tsv("../../resources/dotplots/hap2_vs_hap1.paf", col_names = FALSE, col_types = cols(.default = "c"))
paf <- paf[, 1:12]                      # keep only the 12 mandatory columns
colnames(paf) <- paf_cols

paf <- paf %>%
  mutate(across(c(qlen, qstart, qend, tlen, tstart, tend, nmatch, alen, mapq), as.numeric),
         pident = nmatch / alen)        # approximate % identity per alignment block

# ---- 2. Filter low alignment blocks ----
paf_f <- paf %>%
  filter(alen >= 1000, mapq >= 1)

# ---- 3. Orient reverse-strand alignments for correct diagonal direction ----
paf_plot <- paf_f %>%
  mutate(
    tstart_plot = if_else(strand == "-", tend, tstart),
    tend_plot   = if_else(strand == "-", tstart, tend)
  )

# ---- 4. Simple whole-genome dotplot ----
ggplot(subset(paf_plot, tname == "scaffold_25" & qname == "scaffold_2")) +
  geom_segment(aes(x = qstart, xend = qend,
                    y = tstart_plot, yend = tend_plot,
                    color = strand),
               linewidth = 0.6) +
  scale_color_manual(values = c("+" = "steelblue", "-" = "firebrick")) +
  geom_hline(yintercept = 10964369, lty = 1, color = "pink") +
  geom_hline(yintercept = 12046688, lty = 1, color = "pink") +
  annotate("point", x = (258876087+258880517)/2, y = (10964369+12046688)/2, color = "grey", size = 0.5, shape = 17) +
  #geom_vline(xintercept = , lty = 2, color = "grey") +
  #geom_vline(xintercept = , lty = 2, color = "grey") +
  labs(x = "White (i.e. Haplotype 2), scaffold 2 (bp)", y = "Pink (i.e. Haplotype 1), scaffold 25 (bp)",
       title = "minimap2 white vs pink alignment\nMYB cluster delinated with pink lines",
       color = "Strand") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())

ggsave(paste0("../../results/by-date/", today, "/dotplot_hap2_scaffold2_vs_hap1_scaffold_25.pdf"), width = 6, height = 6)


```
## haplotype 1 vs haplotype 1

Given presence/absence variant of MYB cluster, maybe some scaffolds are not correctly incorporated into the largest 18 scaffolds

```{r}
paf_cols <- c("qname", "qlen", "qstart", "qend", "strand",
              "tname", "tlen", "tstart", "tend",
              "nmatch", "alen", "mapq")

paf <- read_tsv("../../resources/dotplots/hap1_vs_hap1.paf", col_names = FALSE, col_types = cols(.default = "c"))
paf <- paf[, 1:12]                      # keep only the 12 mandatory columns
colnames(paf) <- paf_cols

paf <- paf %>%
  mutate(across(c(qlen, qstart, qend, tlen, tstart, tend, nmatch, alen, mapq), as.numeric),
         pident = nmatch / alen)        # approximate % identity per alignment block

# ---- 2. Filter low alignment blocks ----
paf_f <- paf %>%
  filter(alen >= 1000, mapq >= 1)

# ---- 3. Orient reverse-strand alignments for correct diagonal direction ----
paf_plot <- paf_f %>%
  mutate(
    tstart_plot = if_else(strand == "-", tend, tstart),
    tend_plot   = if_else(strand == "-", tstart, tend)
  )

# ---- 4. Simple whole-genome dotplot ----
ggplot(subset(paf_plot, tname == "scaffold_25" & tname != qname)) +
  geom_segment(aes(x = qstart, xend = qend,
                    y = tstart_plot, yend = tend_plot,
                    color = strand),
               linewidth = 0.6) +
  scale_color_manual(values = c("+" = "steelblue", "-" = "firebrick")) +
  geom_hline(yintercept = 10964369) +
  geom_hline(yintercept = 12046688) +
  labs(x = "Query sequence (bp)", y = "Pink (i.e. Haplotype 1), scaffold 25 (bp)",
       title = "minimap2 whole-genome alignment of Pink vs Pink\nMyb cluster delinated with lines",
       color = "Strand") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())

ggsave(paste0("../../results/by-date/", today, "/dotplot_hap2_scaffold2_vs_hap1_scaffold_25.pdf"), width = 6, height = 6)
```

## haplotype 1 and 2

Code and alignment from yaniv

```{r}
x <- read_tsv("../../resources/dotplots/hap1_vs_hap2.1to1.coords", col_names = c("S1","E1" ,"S2","E2","LEN_1","LEN_2","PCT_IDY","LEN_R","LEN_Q","COV_R","COV_Q","REF_TAG","QUERY_TAG"))

make_plot <- function(scaffold){
  queery <- x|>
    filter(REF_TAG == scaffold) |>
    group_by(QUERY_TAG) |>
    summarise(tot = sum(LEN_2))|>
    mutate(prop = tot / sum(tot))|>
    filter(prop > 0.02)|>
    pull(QUERY_TAG)
  ggplot(x|> filter(REF_TAG==scaffold,QUERY_TAG %in% queery),
         aes(x = S1, xend = E1, y = S2, yend = E2 , color = QUERY_TAG))+
    geom_segment(linewidth = 2)+
    geom_vline(xintercept = 57821464) + 
    geom_vline(xintercept = 57828110) + 
    theme_minimal()+
    theme(axis.line = element_line(color = "black"))+
    ggtitle(scaffold)
}

make_plot("scaffold_1")
ggsave(paste0("../../results/by-date/", today, "/dotplot_scaffold1_g8317.pdf"))

scaffolds <- paste0("scaffold_", 1:36)

pdf(
  paste0("../../results/by-date/", today, "/hap1_vs_hap2_scaffolds_1_to_18.pdf"),
  width = 10,
  height = 7
)

purrr::walk(scaffolds, \(scaffold) {
  print(make_plot(scaffold))
})

dev.off()
```

# myb orthogroup gene tree

```{r}
# Read the tree
my_tree <- read.tree("../OG0001281_with_g166640.tree")

# Basic plot
plot(my_tree, 
     type = "phylogram")
```

# mapping bias

```{r}
idxstats <- list.files(path = "../results/vg/hap1_vs_hap2", pattern = "idxstats.txt", recursive = T, full.names = T)

# Read and bind them while keeping the file names
idxstats <- idxstats %>% 
  set_names() %>% 
    map_dfr(fread, .id = "source_file")

# calculate 
idxstats <- idxstats %>%
  group_by(source_file) %>%
  reframe(mapped = sum(V3),
            unmapped = sum(V4)) %>%
  summarize(sample = 1:8,
            group = sample %% 2,
            pct_mapped = mapped/(mapped+unmapped))

wilcox.test()
```

# genome scans: pi, fst, tajima's D

```{r}
div_scans <- list.files(path = "../results/vg/hap1_vs_hap2", pattern = "pi_windows_diversity", recursive = T, full.names = T)

# Read and bind them while keeping the file names
div_scans <- div_scans %>% 
  set_names() %>% 
  map_dfr(~fread(.x, col.names = c("chrom","start","end","total.masked","total.missing","total.empty","total.numeric","total.invariant","total.passed","missing","numeric","passed","theta_pi","theta_watterson","tajimas_d")), .id = "source_file")

# separate out sample and site class
div_scans$sample <- gsub("../results/vg/hap1_vs_hap2/|/diversity.*", "", div_scans$source_file)

div_scans$site_class <- gsub("../results/vg/hap1_vs_hap2/.*/diversity/|/pi_windows.*", "", div_scans$source_file)

head(div_scans)

#### plot total diversity ####
cds_scans <- subset(div_scans, site_class == "cds")

# 1. Split the data frame into a list of data frames by your variable
data_list <- split(cds_scans, list(cds_scans$chrom, cds_scans$sample), drop = T)

# 2. Open the PDF device
pdf(paste0("../../results/by-date/", today, "/nucleotide_diversity_scans.pdf"), width = 8, height = 6)

# 3. Map over the list, generate a standard plot for each subset, and print
walk(data_list, function(sub_data) {
p <- ggplot(sub_data,  aes(x = (start + end)/2, y = theta_pi)) +
  geom_point(shape = 16) +
  geom_smooth(method = "loess", span = 0.2, se = FALSE) +
  labs(x = "Window midpoint (bp)", 
       y = "Nucleotide diversity", 
       title = paste0(unique(sub_data$chrom), ", Sample: ", unique(sub_data$sample)))
  
  print(p)
})

# 4. Close the PDF device
dev.off()

#### plot tajima's D ####

# 2. Open the PDF device
pdf(paste0("../../results/by-date/", today, "/tajimas_d_scans.pdf"), width = 8, height = 6)

# 3. Map over the list, generate a standard plot for each subset, and print
walk(data_list, function(sub_data) {
p <- ggplot(sub_data,  aes(x = (start + end)/2, y = tajimas_d)) +
  geom_point(shape = 16) +
  geom_smooth(method = "loess", span = 0.2, se = FALSE) +
  labs(x = "Window midpoint (bp)", 
       y = "Tajima's D", 
       title = paste0(unique(sub_data$chrom), 
                      ", Sample: ", 
                      unique(sub_data$sample)))
  
  print(p)
})

# 4. Close the PDF device
dev.off()

#### plot piN/piS ####

```
# differentiation matrix

```{r}
dxy4 <- fread("../results/vg/hap2_vs_hap1/fst/four/grenedalf_results_genome_pi-between.csv")

pi4 <- list.files(path = "../results/vg/hap2_vs_hap1", pattern = "grenedalf_results_genome_diversity.csv", recursive = T, full.names = T)

pi4 <- pi4 %>% 
  set_names() %>% 
  map_dfr(~fread(.x, col.names = c("chrom","start","end","total.masked","total.missing","total.empty","total.numeric","total.invariant","total.passed","missing","numeric","passed","theta_pi","theta_watterson","tajimas_d")), .id = "source_file")

pi4$sample_id <- gsub("/", "", str_extract(pi4$source_file, "/[1-8]/"))

pi4$site_class <- gsub("diversity/|/grenedalf", "", str_extract(pi4$source_file, "diversity/.*/grenedalf"))

pi4 <- pi4 %>% filter(site_class == "four")

# reformat grenedalf table into matrix
pi_long <- dxy4 %>%
  select(matches("^\\d+:\\d+\\.pi-between$")) %>%
  pivot_longer(everything(), names_to = "pair", values_to = "pi") %>%
  mutate(pair = str_remove(pair, "\\.pi-between$")) %>%
  separate(pair, into = c("i", "j"), sep = ":", convert = TRUE)

n <- max(pi_long$i, pi_long$j)

pi_mat <- matrix(NA_real_, n, n, dimnames = list(1:n, 1:n))
pi_mat[cbind(pi_long$i, pi_long$j)] <- pi_long$pi
pi_mat[cbind(pi_long$j, pi_long$i)] <- pi_long$pi

diag(pi_mat) <- pi4$theta_pi

rownames(pi_mat) = c("1P", "1W", "2P", "2W", "3P", "3W", "4P", "4W")

colnames(pi_mat) = c("1P", "1W", "2P", "2W", "3P", "3W", "4P", "4W")

# diversity heatmap
# dxy4/pi-between off diagonal
# pi4 on diagonal
pi_df <- pi_mat %>%
  as.data.frame() %>%
  rownames_to_column("row") %>%
  pivot_longer(-row, names_to = "col", values_to = "pi") %>%
  mutate(
    row   = factor(row, levels = rownames(pi_mat)),
    col   = factor(col, levels = colnames(pi_mat)),
    label = round(pi, 4)
  ) %>%
  filter(as.integer(row) >= as.integer(col))

ggplot(pi_df, aes(x = col, y = row, fill = pi)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = label), size = 3.2) +
  scale_fill_scico(palette = "lajolla", name = "\u03c04", direction = -1, begin = 0.25) +
  scale_y_discrete(limits = rev(levels(pi_df$row))) +
  coord_fixed() +
  theme(
    axis.title = element_blank(),
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

ggsave(paste0("../../results/by-date/", today, "/pi4_dxy4_heatmap.pdf"), device = cairo_pdf, width = 6, height = 6)
ggsave(paste0("../../results/by-date/", today, "/pi4_dxy4_heatmap.png"), width = 6, height = 6)

# fst matrix
fst_mat <- fread("../results/vg/hap2_vs_hap1/fst/four/grenedalf_results_genome_fst-matrix.csv")

fst_mat$fst = c("1P", "1W", "2P", "2W", "3P", "3W", "4P", "4W")
colnames(fst_mat) = c("fst", "1P", "1W", "2P", "2W", "3P", "3W", "4P", "4W")

fst_df <- fst_mat %>% 
  pivot_longer(-fst, names_to = "col", values_to = "Fst") %>%
  mutate(
    row   = factor(fst, levels = fst_mat$fst),
    col   = factor(col, levels = colnames(fst_mat)),
    label = round(Fst, 3)
  ) %>%
  filter(as.integer(row) >= as.integer(col))

ggplot(fst_df, aes(x = col, y = row, fill = Fst)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = label), size = 3.2) +
  scale_fill_scico(palette = "buda", name = "Fst", direction = -1, begin = 0.1) +
  scale_y_discrete(limits = rev(levels(pi_df$row))) +
  coord_fixed() +
  theme(
    axis.title = element_blank(),
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

ggsave(paste0("../../results/by-date/", today, "/fst_heatmap.pdf"), device = cairo_pdf, height = 6, width = 6)
ggsave(paste0("../../results/by-date/", today, "/fst_heatmap.png"), height = 6, width = 6)
```

# coverage from mpileup

```{r}
mpileup_files <- list.files(path = "../results/vg/hap2_vs_hap1", pattern = "\\.mpileup$", recursive = T, full.names = T)

mpileup <- mpileup_files %>%
  set_names() %>%
  map_dfr(~fread(.x, col.names = c("chrom", "pos", "ref", "depth", "bases", "quals")), .id = "source_file")

mpileup$sample_id <- gsub("/", "", str_extract(mpileup$source_file, "/[1-8]/"))

# coverage scan along the scaffold
ggplot(mpileup, aes(x = pos, y = depth, color = sample_id)) +
  geom_line() +
  labs(x = "Position on hap1_scaffold_25 (bp)",
       y = "Read depth",
       color = "Sample",
       title = "Coverage along hap1_scaffold_25")

ggsave(paste0("../../results/by-date/", today, "/mpileup_coverage_scan.pdf"), device = cairo_pdf, width = 8, height = 6)
ggsave(paste0("../../results/by-date/", today, "/mpileup_coverage_scan.png"), width = 8, height = 6)

# depth distribution per sample
ggplot(mpileup, aes(x = depth, fill = sample_id)) +
  geom_histogram(bins = 50) +
  facet_wrap(~sample_id, scales = "free_y") +
  labs(x = "Read depth",
       y = "Number of positions",
       title = "Read depth distribution on hap1_scaffold_25") +
  theme(legend.position = "none")

ggsave(paste0("../../results/by-date/", today, "/mpileup_depth_histogram.pdf"), device = cairo_pdf, width = 8, height = 6)
ggsave(paste0("../../results/by-date/", today, "/mpileup_depth_histogram.png"), width = 8, height = 6)
```