---
title: "Investigate pink/white polymorphism in Leptosiphon parviflorus"
output: html_notebook
---

# clean slate

```{r}
rm(list = ls())
gc()
.rs.restartR(clean = TRUE)
```

```{r}
library(ggplot2)
library(tidyverse)
library(data.table)
library(ggforce)
today <- Sys.Date()
```

# global ggplot2 theme
```{r}
theme_set(
  theme_minimal() +
    theme(
      text = element_text(
        size = 16,
        family = "Helvetica",
        colour = "black"
      ),
      axis.text = element_text(colour = "black"),
      plot.title = element_text(hjust = 0.5),
      strip.text = element_text(colour = "black"),
      panel.grid.minor = element_blank()
    )
)
```

# sift interproscan output

```{r}

part1 <- fread("../../resources/interproscan/iprscan5-R20260721-003940-0554-50156080-p1m.tsv")

part2 <- fread("../../resources/interproscan/iprscan5-R20260721-003958-0335-42354918-p2m.tsv")

interproscan_results <- rbind(part1, part2)

panther_results <- interproscan_results %>% 
  filter(V4 == "PANTHER") %>%
  filter(!str_detect(V6, "POLYMERASE")) %>%
  filter(!str_detect(V6, "TRANSPOSASE")) %>%
  filter(!str_detect(V6, "DUF"))

```

# dotplots

## haplotype 1 vs primary

```{r}
# ---- 1. Read PAF ----
paf_cols <- c("qname", "qlen", "qstart", "qend", "strand",
              "tname", "tlen", "tstart", "tend",
              "nmatch", "alen", "mapq")

paf <- read_tsv("../../resources/dotplots/hap1_vs_primary.paf", col_names = FALSE, col_types = cols(.default = "c"))
paf <- paf[, 1:12]                      # keep only the 12 mandatory columns
colnames(paf) <- paf_cols

paf <- paf %>%
  mutate(across(c(qlen, qstart, qend, tlen, tstart, tend, nmatch, alen, mapq), as.numeric),
         pident = nmatch / alen)        # approximate % identity per alignment block

# ---- 2. Filter low alignment blocks ----
paf_f <- paf %>%
  filter(alen >= 1000, mapq >= 1)

# ---- 3. Orient reverse-strand alignments for correct diagonal direction ----
paf_plot <- paf_f %>%
  mutate(
    tstart_plot = if_else(strand == "-", tend, tstart),
    tend_plot   = if_else(strand == "-", tstart, tend)
  )

# ---- 4. Simple whole-genome dotplot ----
ggplot(paf_plot) +
  geom_segment(aes(x = qstart, xend = qend,
                    y = tstart_plot, yend = tend_plot,
                    color = strand),
               linewidth = 0.6) +
  scale_color_manual(values = c("+" = "steelblue", "-" = "firebrick")) +
  labs(x = "Query (bp)", y = "Target (bp)",
       title = "minimap2 whole-genome alignment",
       color = "Strand") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())

## 5. specific scaffold
for(tscaf in paste0("scaffold_", 1:51)){
    
  ggplot(subset(paf_plot, qname == "scaffold_25" & tname == tscaf)) +
  geom_segment(aes(x = qstart, xend = qend,
                    y = tstart_plot, yend = tend_plot,
                    color = strand),
               linewidth = 0.6) +
  scale_color_manual(values = c("+" = "steelblue", "-" = "firebrick")) +
  labs(x ="Haplotype 1, scaffold 25 (bp)", y = paste0("Primary ", tscaf, " (bp)"),
       title = "minimap2 whole-genome alignment",
       color = "Strand") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
  
  ggsave(paste0("../../results/by-date/", today, "/dotplot_hap1_scaffold_25_vs_primary_scaffold_", tscaf, ".pdf"), width = 6, height = 6)
}

```

## haplotype 1 and haplotype 2

check if myb cluster also present on haplotype 2

```{r}
paf_cols <- c("qname", "qlen", "qstart", "qend", "strand",
              "tname", "tlen", "tstart", "tend",
              "nmatch", "alen", "mapq")

paf <- read_tsv("../../resources/dotplots/hap2_vs_hap1.paf", col_names = FALSE, col_types = cols(.default = "c"))
paf <- paf[, 1:12]                      # keep only the 12 mandatory columns
colnames(paf) <- paf_cols

paf <- paf %>%
  mutate(across(c(qlen, qstart, qend, tlen, tstart, tend, nmatch, alen, mapq), as.numeric),
         pident = nmatch / alen)        # approximate % identity per alignment block

# ---- 2. Filter low alignment blocks ----
paf_f <- paf %>%
  filter(alen >= 1000, mapq >= 1)

# ---- 3. Orient reverse-strand alignments for correct diagonal direction ----
paf_plot <- paf_f %>%
  mutate(
    tstart_plot = if_else(strand == "-", tend, tstart),
    tend_plot   = if_else(strand == "-", tstart, tend)
  )

# ---- 4. Simple whole-genome dotplot ----
ggplot(subset(paf_plot, tname == "scaffold_25" & qname == "scaffold_2")) +
  geom_segment(aes(x = qstart, xend = qend,
                    y = tstart_plot, yend = tend_plot,
                    color = strand),
               linewidth = 0.6) +
  scale_color_manual(values = c("+" = "steelblue", "-" = "firebrick")) +
  geom_hline(yintercept = 10964369, lty = 1, color = "pink") +
  geom_hline(yintercept = 12046688, lty = 1, color = "pink") +
  annotate("point", x = (258876087+258880517)/2, y = (10964369+12046688)/2, color = "grey", size = 0.5, shape = 17) +
  #geom_vline(xintercept = , lty = 2, color = "grey") +
  #geom_vline(xintercept = , lty = 2, color = "grey") +
  labs(x = "White (i.e. Haplotype 2), scaffold 2 (bp)", y = "Pink (i.e. Haplotype 1), scaffold 25 (bp)",
       title = "minimap2 white vs pink alignment\nMYB cluster delinated with pink lines",
       color = "Strand") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())

ggsave(paste0("../../results/by-date/", today, "/dotplot_hap2_scaffold2_vs_hap1_scaffold_25.pdf"), width = 6, height = 6)


```
## haplotype 1 vs haplotype 1

Given presence/absence variant of MYB cluster, maybe some scaffolds are not correctly incorporated into the largest 18 scaffolds

```{r}
paf_cols <- c("qname", "qlen", "qstart", "qend", "strand",
              "tname", "tlen", "tstart", "tend",
              "nmatch", "alen", "mapq")

paf <- read_tsv("../../resources/dotplots/hap1_vs_hap1.paf", col_names = FALSE, col_types = cols(.default = "c"))
paf <- paf[, 1:12]                      # keep only the 12 mandatory columns
colnames(paf) <- paf_cols

paf <- paf %>%
  mutate(across(c(qlen, qstart, qend, tlen, tstart, tend, nmatch, alen, mapq), as.numeric),
         pident = nmatch / alen)        # approximate % identity per alignment block

# ---- 2. Filter low alignment blocks ----
paf_f <- paf %>%
  filter(alen >= 1000, mapq >= 1)

# ---- 3. Orient reverse-strand alignments for correct diagonal direction ----
paf_plot <- paf_f %>%
  mutate(
    tstart_plot = if_else(strand == "-", tend, tstart),
    tend_plot   = if_else(strand == "-", tstart, tend)
  )

# ---- 4. Simple whole-genome dotplot ----
ggplot(subset(paf_plot, tname == "scaffold_25" & tname != qname)) +
  geom_segment(aes(x = qstart, xend = qend,
                    y = tstart_plot, yend = tend_plot,
                    color = strand),
               linewidth = 0.6) +
  scale_color_manual(values = c("+" = "steelblue", "-" = "firebrick")) +
  geom_hline(yintercept = 10964369) +
  geom_hline(yintercept = 12046688) +
  labs(x = "Query sequence (bp)", y = "Pink (i.e. Haplotype 1), scaffold 25 (bp)",
       title = "minimap2 whole-genome alignment of Pink vs Pink\nMyb cluster delinated with lines",
       color = "Strand") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())

ggsave(paste0("../../results/by-date/", today, "/dotplot_hap2_scaffold2_vs_hap1_scaffold_25.pdf"), width = 6, height = 6)
```

## haplotype 1 and 2

Code and alignment from yaniv

```{r}
x <- read_tsv("../../resources/dotplots/hap1_vs_hap2.1to1.coords", col_names = c("S1","E1" ,"S2","E2","LEN_1","LEN_2","PCT_IDY","LEN_R","LEN_Q","COV_R","COV_Q","REF_TAG","QUERY_TAG"))

make_plot <- function(scaffold){
  queery <- x|>
    filter(REF_TAG == scaffold) |>
    group_by(QUERY_TAG) |>
    summarise(tot = sum(LEN_2))|>
    mutate(prop = tot / sum(tot))|>
    filter(prop > 0.02)|>
    pull(QUERY_TAG)
  ggplot(x|> filter(REF_TAG==scaffold,QUERY_TAG %in% queery),
         aes(x = S1, xend = E1, y = S2, yend = E2 , color = QUERY_TAG))+
    geom_segment(linewidth = 2)+
    geom_vline(xintercept = 57821464) + 
    geom_vline(xintercept = 57828110) + 
    theme_minimal()+
    theme(axis.line = element_line(color = "black"))+
    ggtitle(scaffold)
}

make_plot("scaffold_1")
ggsave(paste0("../../results/by-date/", today, "/dotplot_scaffold1_g8317.pdf"))

scaffolds <- paste0("scaffold_", 1:36)

pdf(
  paste0("../../results/by-date/", today, "/hap1_vs_hap2_scaffolds_1_to_18.pdf"),
  width = 10,
  height = 7
)

purrr::walk(scaffolds, \(scaffold) {
  print(make_plot(scaffold))
})

dev.off()
```

# genome scans: pi, fst, tajima's D

```{r}
div_scans <- list.files(path = "../results/vg/hap1_vs_hap2", pattern = "pi_windows_diversity", recursive = T, full.names = T)

# Read and bind them while keeping the file names
div_scans <- div_scans %>% 
  set_names() %>% 
  map_dfr(fread, .id = "source_file")

# separate out sample and site class
div_scans$sample <- gsub("../results/vg/hap1_vs_hap2/|/diversity.*", "", div_scans$source_file)

div_scans$site_class <- gsub("../results/vg/hap1_vs_hap2/.*/diversity/|/pi_windows.*", "", div_scans$source_file)

head(div_scans)

#### plot total diversity ####

cds_scans <- subset(div_scans, site_class == "cds")

# 1. Split the data frame into a list of data frames by your variable
data_list <- split(cds_scans, list(cds_scans$chrom, cds_scans$sample), drop = T)

# 2. Open the PDF device
pdf(paste0("../../results/by-date/", today, "/nucleotide_diversity_scans.pdf"), width = 8, height = 6)

# 3. Map over the list, generate a standard plot for each subset, and print
walk(data_list, function(sub_data) {
p <- ggplot(sub_data,  aes(x = (start + end)/2, y = mapped.reordered.theta_pi)) +
  geom_point(shape = 16) +
  geom_smooth(method = "loess", span = 0.2, se = FALSE) +
  labs(x = "Window midpoint (bp)", 
       y = "Nucleotide diversity", 
       title = paste0(unique(sub_data$chrom), ", Sample: ", unique(sub_data$sample)))
  
  print(p)
})

# 4. Close the PDF device
dev.off()

#### plot tajima's D ####

# 2. Open the PDF device
pdf(paste0("../../results/by-date/", today, "/tajimas_d_scans.pdf"), width = 8, height = 6)

# 3. Map over the list, generate a standard plot for each subset, and print
walk(data_list, function(sub_data) {
p <- ggplot(sub_data,  aes(x = (start + end)/2, y = mapped.reordered.tajimas_d)) +
  geom_point(shape = 16) +
  geom_smooth(method = "loess") +
  geom_vline(xintercept = 258876087) +
  geom_vline(xintercept = 258880517) +
  labs(x = "Window midpoint (bp)", 
       y = "Tajima's D", 
       title = paste0(unique(sub_data$chrom), ", Sample: ", unique(sub_data$sample)))
  
  print(p)
})

# 4. Close the PDF device
dev.off()

#### plot piN/piS ####

```
